Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-08 paper/paper.tex
further revision to abstract and intro

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which include the size of the maximum independent sets, the number and enumeration of independent sets of a given size.
% which include the size of the maximum independent sets, the number of independent sets and maximal independent sets of a given size, and enumeration of independent sets and maximal independent sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the tree width of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the overlap gap property on three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting and hard computational problems concerning various properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$. 
One of the independent set properties that many computational complexity theorists are most interested in is to find a maximum independent set (MIS) and the size of the MIS, $\alpha(G) \equiv \max_{I}|I|$,
not only because it has a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
but also because it is a well-known NP-complete problem~\cite{Hastad1996} that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.
Some other properties, for example, the number and enumeration of independent sets or maximal independent sets of a given size are euqally important in many practical applications, however, are not studied enough.
Some of them are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022} \red{replace with experiment paper when ready};
they can, for example, be used to understand phase transitions, to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
%\blue{The primary interest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.}
%Finding $\alpha(G)$ exactly is a hard computational problem, and it is NP-hard even to approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
%meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
%Naive exhaustive search for an MIS requires a computational time of $O\left(2^{|V|} \right)$.
%There is immense interest in finding better algorithms to solve the MIS problem, not only because it has 
%a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
%but also because it is a well-known NP-complete problem that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.

@@

In this paper, we focus on computing various independent set \textit{properties}. % We introduce a generic programming tensor-network framework to compute the various \textit{properties} pertaining to independent sets. 

@@

We map the computation of these independent set properties into generic tensor network contraction with specially designed tensor element algebra.

@@

The complexity of our tensor-network based algorithms is similar to that of dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} that dedicated to solving the MIS size,
which scales exponentially with the tree width, ${\rm tw}(G)$, of the graph.
For sparse graphs, this quantity is usually much smaller than the number of vertices~\cite{Fomin2006}.
This indicates it is a method that works better for sparse graphs,
unlike another standard approach for solving the MIS size, the branching algorithm ~\cite{Robson1986,Tarjan1977,Xiao2017}, which usually works better for dense graphs.

@@

We benchmark our algorithms in \Sec{sec:benchmark} by computing various independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performacne.
The high performance of our algorithms benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021},

@@

Lastly, we provide a few examples and demonstrate the versatility of our tool by computing the entropy constant for some hardcore lattice gases on 2D square lattices and analyzing the overlap gap property on three regular graphs.

@@

\section{Benchmarks and case studies}\label{sec:benchmark}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small tree width that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.

@@

Figure (a) shows the time and space complexity of tensor network contraction for variety of graph sizes, where the space complexity is the same as the tree width of the problem graph.
The contraction order is obtained using the local search algorithm in Ref.~\cite{Kalachev2021}.
In practice, slicing technique is used for graphs with tree width greater than 27 to fit the computation into a 16GB memory.
One can see all the computing times in figure (b), (c) and (d) have a strong correlation with the tree width.
%The computations shown in \Fig{fig:benchmark} include independent set counting, finding maximum independent set size (panel b),
%computing independence polynomial (panel c) and find one independent set and enumerating independent sets (panel d).
Among these benchmarks, computational tasks with data types \texttt{T (CPU)},
\texttt{$\mathbbm{R}$ (CPU)}, \texttt{$\mathbbm{R}$ (GPU)}, \texttt{$\mathbbm{C}$ (CPU)}, \texttt{$\mathbbm{C}$ (GPU)} and \texttt{T+bounding (CPU)} can utilize fast BLAS functions to evaluate tensor contractions, hence 
are much faster comparing to non-BLAS element types in the same category.
GPU computes much faster than CPU in all cases when the problem scale is large enough such that the actual computing time is comparable or larger than the launching overhead of CUDA kernels.
Most algebras can be computed on GPU, except those requiring dynamic sized structures, i.e. \texttt{PN} and \texttt{SN}.
In figure (c), one can see the Fourier transformation based method is the fastest in computing the independence polynomial,
however, it maybe suffer from the round off errors. The finite field (GF$(p)$) approach is the only method that does not have round-off errors and can run on a GPU.
In figure (d), one can see the technique to bound the enumeration space improves the performance for more than one order in enumerating the MISs.
Bounding can also reduce the memory usage significantly, without which the largest computable graph size is only $\sim150$ on a device with 32GB main memory.

@@

%Dynamic programming can solve monadic second order logic (MSO) definable problem on a graph with small tree width efficiently~\cite{Courcelle1990,Barr2020}.

@@

One can see that the average degree of a tensor is increased.
The computational complexity of this new tensor network contraction is often greater than the one for computing the independence polynomial.
However, for most sparse graphs, this tensor network contraction approach is still much faster than enumerating all the maximal cliques on its complement graph using the Bron-Kerbosch algorithm~\cite{Bron1973}, which is the standard algorithm that we are aware of to compute the maximal independence polynomial.
We show the benchmark of computing the maximal independent set properties in \Fig{fig:benchmark-maximal},
including a comparison to the Bron-Kerbosch algorithm from Julia package Graphs~\cite{Graphs}.
%Since the network for computing maximal independent set properties is different from the one for computing the independence polynomial,
the tree width of this tensor network is significantly larger, hence only small graphs can be benchmarked.
The time for the tensor network approach and the Bron-Kerbosch approach to enumerate all maximal independent sets are comparable,
while the tensor network does counting much more efficiently.
Due to the memory limit, this Bron-Kerbosch algorithm stops working at size $70$ and above.

>>>>

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which include, for example, the size of the maximum independent sets and the number and enumeration of independent sets of a given size.
% which include the size of the maximum independent sets, the number of independent sets and maximal independent sets of a given size, and enumeration of independent sets and maximal independent sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting and hard computational problems concerning various properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$. 
One of the paradigmatic independent set \textit{properties} % that many computational complexity theorists are most interested in 
is the maximum independent set (MIS) and the size of the MIS, $\alpha(G) \equiv \max_{I}|I|$. 
There is widespread interest in this MIS problem not only because it is a well-known NP-complete problem~\cite{Hastad1996} that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011},
but also because it has a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
Some other properties, such as the number and enumeration of independent sets or maximal independent sets of a given size, are equally interesting, if not more important, in many practical applications. These other properties can be found in physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022} \red{replace with experiment paper when ready};
they can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007}, to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}. However, there is a lack of general and versatile tool to compute the different properties of independent sets.
%\blue{The primary interest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.}
%Finding $\alpha(G)$ exactly is a hard computational problem, and it is NP-hard even to approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
%meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
%Naive exhaustive search for an MIS requires a computational time of $O\left(2^{|V|} \right)$.
%There is immense interest in finding better algorithms to solve the MIS problem, not only because it has 
%a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
%but also because it is a well-known NP-complete problem that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.

@@

% We introduce a generic programming tensor-network framework to compute the various \textit{properties} pertaining to independent sets. 

@@

%In this paper, we focus on computing various independent set \textit{properties}. 
In this paper, we introduce a generic programming tensor-network framework to compute a number of different independent set properties.
We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.

@@

Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
The complexity of our tensor-network based algorithms is similar to that of dynamic programming algorithms~\cite{Courcelle1990, Fomin2013}, which have a complexity scaling exponentially with the treewidth, ${\rm tw}(G)$, of the graph;
for sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.
% Thus, we expect our algorithms to works well for sparse graphs,
% which is complementary to another standard approach for solving the MIS size, the branching algorithm ~\cite{Robson1986, Tarjan1977, Xiao2017}, which usually works better for dense graphs.

@@

We benchmark our algorithms in \Sec{sec:benchmark} by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
The high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021},

@@

Lastly, we provide a few examples and demonstrate the versatility of our tool by computing the entropy constant for some hardcore lattice gases on 2D square lattices and analyzing the presence of overlap gap property on three regular graphs.

@@

\section{Benchmarks and case studies}\label{sec:benchmark}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small treewidth that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.

@@

Figure (a) shows the time and space complexity of tensor network contraction for variety of graph sizes, where the space complexity is the same as the treewidth of the problem graph.
The contraction order is obtained using the local search algorithm in Ref.~\cite{Kalachev2021}.
In practice, slicing technique is used for graphs with treewidth greater than 27 to fit the computation into a 16GB memory.
One can see all the computing times in figure (b), (c) and (d) have a strong correlation with the treewidth.
%The computations shown in \Fig{fig:benchmark} include independent set counting, finding maximum independent set size (panel b),
%computing independence polynomial (panel c) and find one independent set and enumerating independent sets (panel d).
Among these benchmarks, computational tasks with data types \texttt{T (CPU)},
\texttt{$\mathbbm{R}$ (CPU)}, \texttt{$\mathbbm{R}$ (GPU)}, \texttt{$\mathbbm{C}$ (CPU)}, \texttt{$\mathbbm{C}$ (GPU)} and \texttt{T+bounding (CPU)} can utilize fast BLAS functions to evaluate tensor contractions, hence 
are much faster comparing to non-BLAS element types in the same category.
GPU computes much faster than CPU in all cases when the problem scale is large enough such that the actual computing time is comparable or larger than the launching overhead of CUDA kernels.
Most algebras can be computed on GPU, except those requiring dynamic sized structures, i.e. \texttt{PN} and \texttt{SN}.
In figure (c), one can see the Fourier transformation based method is the fastest in computing the independence polynomial,
however, it maybe suffer from the round off errors. The finite field (GF$(p)$) approach is the only method that does not have round-off errors and can run on a GPU.
In figure (d), one can see the technique to bound the enumeration space improves the performance for more than one order in enumerating the MISs.
Bounding can also reduce the memory usage significantly, without which the largest computable graph size is only $\sim150$ on a device with 32GB main memory.

@@

%Dynamic programming can solve monadic second order logic (MSO) definable problem on a graph with small treewidth efficiently~\cite{Courcelle1990,Barr2020}.

@@

One can see that the average degree of a tensor is increased.
The computational complexity of this new tensor network contraction is often greater than the one for computing the independence polynomial.
However, for most sparse graphs, this tensor network contraction approach is still much faster than enumerating all the maximal cliques on its complement graph using the Bron-Kerbosch algorithm~\cite{Bron1973}, which is the standard algorithm that we are aware of to compute the maximal independence polynomial.
We show the benchmark of computing the maximal independent set properties in \Fig{fig:benchmark-maximal},
including a comparison to the Bron-Kerbosch algorithm from Julia package Graphs~\cite{Graphs}.
%Since the network for computing maximal independent set properties is different from the one for computing the independence polynomial,
the treewidth of this tensor network is significantly larger, hence only small graphs can be benchmarked.
The time for the tensor network approach and the Bron-Kerbosch approach to enumerate all maximal independent sets are comparable,
while the tensor network does counting much more efficiently.
Due to the memory limit, this Bron-Kerbosch algorithm stops working at size $70$ and above.
